---
title: "Did Republican or Democrat voters experience more difficulty voting in the 2020 election?"
subtitle: "DATASCI 203: Lab 1 Hypothesis Testing"
author: "Nicolas Loffreda, Ethan Duncan, Jeremy Lan" 
output:
  bookdown::pdf_document2: 
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage
\setcounter{page}{1}

```{r load packages and set options, include=FALSE}
library(tidyverse) 
library(magrittr)
library(knitr)
library(patchwork)
library(moments)
library(dplyr)
library(readr)
library(ggplot2)
require(effsize)
library(effsize)
theme_set(theme_bw())
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo=FALSE, message=FALSE)
```
```
```{r, warning=FALSE, message=FALSE}

anes <- read_csv("anes_timeseries_2020_csv_20220210.csv")
# Keep only variables of interest
keeps = c(
  "V201600",   # gender
  "V201507x",  # age
  "V201511x",  # education
  "V201549x",  # race
  "V201567",   # children
  "V201617x",  # family income
  "V202109x",  # turnout in 2020 election
  "V202110x",  # 2020 presi vote
  "V202111x",  # 2020 house vote
  "V202112x",  # 2020 senate vote
  "V202113x",  # 2020 gov vote
  "V201231x",  # party ID
  "V201019",   # pre. intend to register to vote
  "V202051",   # post. registered to vote
  "V202117",   # voting method
  "V202118",   # usual voting method
  "V202119",   # how difficult to vote
  "V202120a",  # registration issues?
  "V202120b",  # concern on ID card?
  "V202120c",  # diffculties with absentee ballot
  "V202120d",  # confusion on ballot/machine
  "V202120e",  # difficulty getting to polling place
  "V202120f",  # long wait time at polling place
  "V202120g",  # work schedule
  "V202120h",  # bad weather
  "V202120i",  # issue mailing ballot
  "V202120j",  # other issue
  "V202120k",  # no issues voting
  "V202121",   # how long the wait
  "V202122",   # how long to get to polling line
  "V202123",   # main reason not to vote
  "V202124"    # other reason not to vote
)
column_names = c(
  "gender",
  "age",
  "education",
  "race",
  "children",
  "income",
  "voted_2020",
  "v2020_presi",
  "v2020_house",
  "v2020_senate",
  "v2020_gov",
  "party",
  "pre_intent_to_vote",
  "post_registered_to_tove",
  "voting_method",
  "usual_voting_method",
  "diffvote",
  "diffvote_reg",
  "diffvote_idcard",
  "diffvote_aballot",
  "diffvote_confmach",
  "diffvote_gopolling",
  "diffvote_wait",
  "diffvote_work",
  "diffvote_weather",
  "diffvote_mail",
  "diffvote_other",
  "diffvote_none",
  "diffvote_wait_time",
  "diffvote_gopolling_time",
  "diffvote_mainreason",
  "diffvote_otherreason"
)
anes_filt = anes[keeps]
colnames(anes_filt) = column_names

anes_filt = anes_filt %>%
  mutate(party_self = case_when(
    party %in% c(1, 2, 3) ~ "Democrat",
    party %in% c(5, 6, 7) ~ "Republican",
    party %in% c(4)       ~ "Independent"
  ))
anes_voted = anes_filt %>% filter(voted_2020 == 1)
anes_voted_party_clean = anes_voted %>% 
  filter(
    # Filter out NAs or negative values in difficulty voting
    diffvote >= 0 & !is.na(diffvote) &
    # Keep only reps and demos
    party_self %in% c("Republican", "Democrat"))
```

```
# Importance and Context

Voter turnout is essential to a functioning democracy, and although not every American experiences difficulties in the voting process, many circumstances can interfere with one's ability to vote. Some barriers are more systemic and tied to the voter registration process, such as difficulties obtaining a ballet or proving residency in a certain district. Other obstacles may present themselves on Election Day itself, such as lack of accessibility to a polling location or inability to find time in the work schedule to submit a ballot. During the 2020 election, the COVID-19 pandemic may have also led more Americans to vote by mail - surveys have indicated that 46\% of 2020 voters voted by mail or absentee compared to 21\% of voters in the 2016 election \footnote{FiveThirtyEight. "What Absentee Voting Looked Like In All 50 States." (2020).}. This rise in mail-in voting may have presented its own challenges as many voters navigated the process for the first time.

In this analysis, we aim to investigate sources of voting difficulty that affected American voters in the 2020 election. Specifically, we hope to address the following research question with a statistically sound approach:

\begin{quote}
  \textit{Did Republican or Democrat voters experience more difficulties voting in the 2020 election?}
\end{quote}

Observing if voting difficulty may be a partisan issue can also shed some insights into opportunities to make elections more accessible to all Americans. Of course, people's voting experiences are affected by much more than just their party affiliation, so we must be careful to not draw any unjustified causal conclusions from the results of our analysis. However, the observations made can lead to additional investigations about how a variety of other socioeconomic factors, in conjunction with political values or party affiliation, might cause obstacles in the voting process.

# Data and Methodology

To answer this research question, we are utilizing data from the 2020 Time Series Study, hosted by the American National Election Studies (ANES). This dataset is nationally representative since ANES conducted a fresh cross-sectional sample that was a random draw from the USPS computerized delivery sequence file, with residential addresses across the 50 states and Washington DC having equal probability of selection \footnote{ANES 2020 Time Series Study: User Guide and Codebook - Page 8K (2022)}. Respondents to the study were asked a variety of questions centered around various election-related and socioeconomic topics. The respondents to the 2020 Time Series Study are drawn from two sample compositions for a total of 8,280 individuals:

\begin{itemize}
\item[--] 5,441 interviews were drawn from a cross-section of all US eligible voters
\item[--] 2,839 interviews were performed on previous respondents of the 2016 ANES Time Series Study
\end{itemize}

We note a few definitions for terms in this study so that others may understand the limitations of our results. We categorized respondents based on their self-reported party affiliation. It is important to note that individuals had the option to respond with "Independent-Democrat" and "Independent-Republican" to reflect more moderate views; we did not exclude these responses, as studies have shown that many Americans may de-emphasize their partisanship as a matter of self-presentation\footnote{Petrocik: "Measuring party support: Leaners are not independents." (2009)}. We excluded respondents that declined to answer this question or that answered "Independent" with no party leaning.

We also exclude participants who did not vote in the 2020 election. Noting these definitions, after cleaning the data based on these parameters, we were left with 3,160 Democrat and 2,714 Republican (5,874 total) respondents who voted in the 2020 election.

The survey also include specific questions concerning voting accessibility and any potential issues voters faced in the voting process. For the main question posed to respondents on this topic "How difficult was it to vote?", data was reported on a 5-point Likert scale ranging from 1 ("Not difficult at all") to 5 ("Extremely difficult"). Data from follow-up questions concerning specific problems (voter registration, work schedule, mailing issues etc.), were reported with binary variables (0 and 1, with 1 recorded if the specific problem was mentioned by the respondent). Excluding those who did not respond to this question, our final sample consists of 3,128 Democrat and 2,700 Republican (5,828 total) respondents.

Preliminary visualization of the distribution of how respondents reported voting difficulty is shown below (Figure 1). Overall, 635 of the 5,828 respondents in our cleaned dataset reported at least some difficulty voting. 

```{r}
anes_voted_party_clean = anes_voted_party_clean %>%
  mutate(diffvote_str = case_when(
    diffvote ==1 ~ "Not Difficult",
    diffvote ==2 ~ "A little difficult",
    diffvote ==3 ~ "Moderately difficult",
    diffvote ==4 ~ "Very difficult",
    diffvote ==5 ~ "Extremely difficult",
  ))

figure1 <- 
  anes_voted_party_clean %>% 
  ggplot() +
  aes(x=reorder(diffvote_str, diffvote), fill=party_self) +
  geom_bar(alpha=0.5, position="dodge", width=0.5) +
  labs(title="Difficulty Voting in 2020 Election: By Party Affiliation", x = "Difficulty Voting", y = "Count of Responses")+
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(size = 8))+
  scale_fill_manual(values=c("blue", "red"), name = "Party Affiliation")
```

```{r plots, fig.cap='Democrat and Republican responses to voting difficulty', fig.pos='!b', fig.height = 3.5}
figure1
```

To test our research question, we proceeded with a Wilcoxon rank-sum test, which is a nonparametric test well-suited for comparing outcomes between two independent groups. As we are dealing with ordinal data (voting difficulty was captured on a Likert scale), specifically the hypothesis of comparisons version of the Wilcoxon rank-sum test will be used. 

The null hypothesis of our Wilcoxon rank-sum test will be stated as follows: 
\begin{quote}
    \textbf{Null Hypothesis:} \textit{The probability that a draw from the Republican voter pool ranks higher than a draw from the Democrat voter pool is the same as the probability that a draw from the Democrat voter pool ranks higher than a draw from the Republican voter pool.}
\end{quote}

To ensure that the Wilcoxon rank-sum test is suitable for our purposes and our data, we must verify the following assumptions: data is independent and identically distributed (IID) data is on an ordinal scale (for the hypothesis of comparisons version of the test).

\begin{itemize}
\item[--] Ordinal data: Likert scale is ordinal. Although the items have a clear rank order, they are not evenly distributed and it would not make sense to conduct arithmetic operations on the data. 
\item[--] IID: The ANES study uses a nationally representative sample of individuals across the USPS computerized delivery sequence file. In theory, each residential address had an equal probability of selection. We acknowledge potential response bias as respondents were offered money for their participation. Additionally, there is opportunity for geographical clustering of respondents, where people who live nearby may have similar voting experiences. However, both of these effects are relatively minimal given the size of the dataset. 
\end{itemize}

\newpage

# Results

The results of the Wilcoxon rank-sum test are shown below. 
```{r echo=T, results='hide'}
wt = wilcox.test(diffvote ~ party_self, data = anes_voted_party_clean, alternative="two.sided")
```

The resultant test statistics (W = 4323774, p = .0035) leads us to reject the null hypothesis at the usual $\alpha$ levels of .05 and .01. Consequently, we conclude that \textbf{the probability that a draw from the Republican voter pool ranks higher than a draw from the Democrat voter pool is not the same as the probability that a draw from the Democrat voter pool ranks higher than a draw from the Republican voter pool}. Although the test is statistically significant, the metric of this test is _not_ very intuitive. Thus, we proceeded to calculate 2 different metrics to interpret the effect size to gauge practical significance: 

1) The correlation effect size for a Wilcoxon rank-sum test $r=\frac{Z}{\sqrt{N}}$, which is very small (0.038):

```{r echo=T, results = 'hide'}
z = qnorm(wt$p.value/2)
N = count(anes_voted_party_clean)
eff_corr = abs(z)/sqrt(N)
as.numeric(eff_corr)
```


2) Cohen's D metric, which specifies how many standard deviations away from the population is one distribution from another one, which is also very small (d estimate = 0.033):

```{r echo=T, results = 'hide'}
cohen.d(diffvote ~ party_self, data = anes_voted_party_clean)
```

All metrics indicate that the test results are not practically significant, despite their statistical significance. The differences in distributions is ultimately quite small. 

# Discussion

This study found there is not a practical difference between difficulty voting for Democrats or difficulty in voting for Republicans in the 2020 election. As stated above, while conducting the Hypothesis of Comparison version of the  Wilcoxon rank-sum test we were led to reject the null hypothesis meaning that there was a statistically significant difference in voting difficulty for Democratic and Republican voters. Although, after using the r-value in the Wilcoxon rank-sum test and the Cohen’s D metric, we found minimal practical significance in the magnitude of our findings. 

We would like to mention several limitations of our test that could affect the conclusions drawn from it.  This study does not get as granular as determining if there are difficulties that disproportionately affect Democratic or Republican voters. Future studies might consider focusing on one voting issue, like physical difficulties, contained in this dataset such as: waiting time to cast the vote, accessibility to voting locations, unclear instructions on how to use the voting mechanism, restrictions due to the COVID-19 pandemic, or other registration reasons. Furthermore, this study does not measure those who didn't vote for any reason; some may have been disinterested in the election process while others may have been so inconvenienced that they were unable to submit their ballots or were disallowed from voting entirely. 

The results of our study have the potential to provide insight to lawmakers and those in authority that have the civic and moral obligation to make the voting process easy, equitable, and accessible to all voters. With future studies focused on individual issues like the ones discussed above, there is the potential to provide valuable insights to political campaigns on both sides of the aisle looking to help potential voters through the voting process. 


